Accelerating Metagenomic Identification of DNA Sequences Using Artificial Neural Networks

### 2025-05-23 article/main.tex
article: include tool name in text

    In conclusion, the study highlights the potential of ANN-based methods to bridge the gap between precision and scalability in genetic taxonomy.
    \todo{ We developed a tool with command-line interface, called Exquisitor that implements presented method.}

@@

    The rest of the paper is organized as follows. Section 2 defines the problem, presents the proposed approach, and provides detailed implementation details \todo{of new tool called Exquisitor}.
    Section 3 describes the experimental setup, including used datasets, the performance metrics for evaluation, and the results obtained from experiments with analysis.
    Finally, section 4 concludes the paper by summarizing the main findings, discuss on the limtiations of proposed approach and suggesting potential directions for future research.

@@

        The core component of \todo{the Exquisitor (proposed solution)} is a processing pipeline designed with flexibility in mind, allowing the integration of various DNA sequence clustering methods. The pipeline consists of a series of steps executed sequentially to perform metagenomic identification. A schematic representation of the pipeline is show in Figure~\ref{Picture:Pipeline}, where input and output data are represented by circles, and each processing step is represented by a rectangle. The pipeline is composed of the following five steps:

@@

    \subsection{\todo{Exquisitor method description}}

@@

        \subsection{Exquisitor implementation details}
            The solution consists of two components. The first is a library that contains components enabling the creation of pipelines for genetic material analysis in the form of metagenomic identification. The second component is a console application, which uses the library's components to allow for the creation and configuration of processing pipelines. Additionally, a web application was created, which allows users to submit analysis tasks through a web browser.

@@

        \subsection{Base-line algorithms}

                \todo{tutaj metody klasyczne, porownawcze. PG: czy trzeba coś tutaj jeszcze dodawać?}

@@

            \subsubsection{$k$-mer embeddings}
                The second approach will be the use of $k$-mers as embeddings, which allows for the representation of DNA sequences as numerical vectors. These vectors will then be compared using Euclidean distance.

        \subsection{Experiments}
            \subsubsection{Metagenomic identification execution time \todo{using Exquisitor}}
                \paragraph{Objective}
                Measuring the execution time of metagenomic identification using different genetic sequence clustering methods.

@@

            \subsubsection{Metagenomic identification quality}
                \paragraph{Objective}
                Examining the quality of metagenomic identification using the implemented methods in comparison to the metagenomic identification of all sequences.

@@

            \subsubsection{Batching sequences}
                Creating a single dissimilarity matrix for very large samples is not optimal, as it significantly impacts the performance of the method. To address this issue, pre-clustering the sequences into batches can provide a solution. Processing in batches may reduce quality but allows for maintaining a short execution time.

@@

    \section{Conclusions}
        \subsection{Summary of Findings}
            The experiments demonstrated that \todo{the proposed method} based on artificial neural networks outperforms traditional approaches in terms of quality, while maintaining execution times comparable to those of the classical $k$-mer method. This highlights the potential of neural networks for improving performance without significant computational overhead.

        \subsection{Future Research Directions}
        Future research could focus on utilizing artificial neural networks as an independent tool for metagenomic identification. Another promising direction is the application of artificial neural networks for dimensionality reduction of genetic sequences, which may improve efficiency and performance in various analyses. \todo{zdanie: Planowane prace wdrożeniowe są skierowane w dostarczeniu interefsejsów do programu Exquisitor, pozwalając na jego integrację z istniejącymi popularnymi potokami (pipeline)}

>>>>

    In conclusion, the study highlights the potential of ANN-based methods to bridge the gap between precision and scalability in genetic taxonomy.
    We developed Exquisitor, a command-line tool implementing proposed method.

@@

    The rest of the paper is organized as follows. Section 2 defines the problem, presents the proposed approach, and provides implementation details of developed tool Exquisitor.
    Section 3 describes the experimental setup, including used datasets, the performance metrics for evaluation, and the results obtained from experiments with analysis.
    Finally, section 4 concludes the paper by summarizing the main findings, discuss on the limtiations of proposed approach and suggesting potential directions for future research.

@@

        The core component of proposed tool, Exquisitor is a processing pipeline designed with flexibility in mind, allowing the integration of various DNA sequence clustering methods. The pipeline consists of a series of steps executed sequentially to perform metagenomic identification. A schematic representation of the pipeline is show in Figure~\ref{Picture:Pipeline}, where input and output data are represented by circles, and each processing step is represented by a rectangle. The pipeline is composed of the following five steps:

@@

    \subsection{Exquisitor: Method Description}

@@

        \subsection{Exquisitor: Implementation Details}
            The solution consists of two components. The first is a library that contains components enabling the creation of pipelines for genetic material analysis in the form of metagenomic identification. The second component is a console application, which uses the library's components to allow for the creation and configuration of processing pipelines. Additionally, a web application was created, which allows users to submit analysis tasks through a web browser.

@@

        \subsection{Base-line Algorithms}

@@

            \subsubsection{$k$-mer Embeddings}
                The second approach will be the use of $k$-mers as embeddings, which allows for the representation of DNA sequences as numerical vectors. These vectors will then be compared using Euclidean distance.

        \subsection{Experiments}
            \subsubsection{Execution Time of Metagenomic Identification with Exquisitor}
                \paragraph{Objective}
                Measuring the execution time of metagenomic identification using different genetic sequence clustering methods.

@@

            \subsubsection{Quality of Metagenomic Identification with Exquisitor}
                \paragraph{Objective}
                Examining the quality of metagenomic identification using the implemented methods in comparison to the metagenomic identification of all sequences.

@@

            \subsubsection{Batching Sequences}
                Creating a single dissimilarity matrix for very large samples is not optimal, as it significantly impacts the performance of the method. To address this issue, pre-clustering the sequences into batches can provide a solution. Processing in batches may reduce quality but allows for maintaining a short execution time.

@@

    \section{Conclusions}
        \subsection{Summary of Findings}
            The experiments demonstrated that the proposed method based on artificial neural networks outperforms traditional approaches in terms of quality, while maintaining execution times comparable to those of the classical $k$-mer method. This highlights the potential of neural networks for improving performance without significant computational overhead.

        \subsection{Future Research Directions}
        Future research could focus on utilizing artificial neural networks as an independent tool for metagenomic identification. Another promising direction is the application of artificial neural networks for dimensionality reduction of genetic sequences, which may improve efficiency and performance in various analyses. Planned development efforts are focused on providing interfaces for the Exquisitor tool, enabling its integration with existing popular metagenomic pipelines.
